---
title: "Posterior sampling with LimoRhyde2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Posterior sampling with LimoRhyde2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  rmarkdown.html_vignette.check_title = FALSE
)
```

## Introduction

In this vignette, we demonstrate how to use LimoRhyde2 to obtain posterior estimates for measurements and rhythmic properties. This is a requirement if you would like to construct credible intervals. The data are microarray samples of mouse liver tissues.

## Setup

We begin by preparing the R environment.

```{r setup}
library(limorhyde2)
library(data.table)
library(ggplot2)
library(qs)
foreach::registerDoSEQ()

theme_set(theme_bw())
```

## Load the data

Next, we load the sample data. The data consist of a gene-by-sample expression matrix and a data.table of sample metadata.

```{r}
exprData = qread(system.file('data', 'GSE34018_data.qs', package = 'limorhyde2')) 
metadata = qread(system.file('data', 'GSE34018_metadata.qs', package = 'limorhyde2'))

#first 8 samples of expression data
head(exprData[, 1:6])
metadata
```

## Fit linear models and get a posterior fit

Next, we fit linear models to the data and obtain a posterior fit. See the [quantifying rhythmicity vignette](quantifying-rhythmicity.html) for more details.

```{r, results = 'hide' }
#fitting initial models
fit = getModelFit(y = exprData, metadata = metadata, condColname = 'cond')

#getting posterior fit
fit = getPosteriorFit(fit)
```

## Sample from the posterior

We use the posterior fit from `getPosteriorFit` to get posterior samples for the coefficient estimates of the model. The default setting is 200 samples. To save time in this vignette, we load the posterior samples for the example dataset from the package data.

```{r, results = 'hide'}
# fit = getPosteriorSamples(fit)
fit = qread(system.file('data', 'GSE34018_fit_ps.qs', package = 'limorhyde2'))
```

## Estimate fitted values

We can also use the posterior samples of coefficients to get posterior samples of expected measurements. Since we set `fitType = 'posterior_samples'`, the resultant data.table will consist of expected measurements for each combination of feature, time-point, and sample. We again limit the features to 3 of the clock genes.

```{r}
clockGeneDt = data.table(id = as.character(c(18626, 12953, 12753)), sym = c('Per1', 'Cry2', 'Clock'))

fittedVals = getExpectedMeas(fit, times = seq(0, 24, by = 0.5), features = clockGeneDt$id,
                             fitType = 'posterior_samples')
head(fittedVals)
```

To get a better understanding of how rhythmicity varies with each posterior sample, we can plot their expected measurements together.

```{r, message=FALSE, fig.align='center', fig.width=8, fig.asp=0.5625}
fittedVals[clockGeneDt, sym := i.sym, on = .(feature = id)]

ggplot(data = fittedVals) +
  geom_line(aes(x = time, y = value, color = cond, group = interaction(cond, posterior_sample)), 
            alpha = 0.2) +
  facet_grid(rows = vars(sym), scales = 'free_y') +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(color = 'Condition', x = 'Circadian time (h)', y = 'Expression (norm.)') +
  scale_color_brewer(palette = 'Dark2')
```

## Get credible intervals for fitted values

Once we've gotten the posterior estimates for the expected measurements, we can use `getExpectedMeasIntervals` to get upper and lower bounds for each combination of feature and timepoint.

```{r}
fittedValsInterval = getExpectedMeasIntervals(fittedVals, method = 'eti')
head(fittedValsInterval)
```

```{r, fig.width=8, fig.asp=0.5625}
#setting up data table with mean fitted values for given time
meanFittedVals = getExpectedMeas(fit, times = seq(0, 24, by = 0.5), features = clockGeneDt$id)
meanFittedVals = merge(meanFittedVals, clockGeneDt, by.x = 'feature', by.y = 'id')

ggplot() +
  geom_ribbon(data = fittedValsInterval, aes(x = time, ymin = lower, ymax = upper,
                                             fill = cond), alpha = 0.3) +
  geom_line(data = meanFittedVals, aes(x = time, y = value, color = cond)) +
  facet_grid(rows = vars(sym), scales = 'free_y') +
  ylab('Expression (norm.)') +
  xlab('Circadian time (h)') +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  scale_fill_brewer(palette = 'Dark2') +
  scale_color_brewer(palette = 'Dark2')
```


## Get rhythmic statistics

We can use the posterior samples of coefficients to get posterior samples of rhythmic statistics. Since we set `fitType = 'posterior samples'`, this step will result in a data.table with rhythmic statistics for each feature in each sample. In this example, we limit the features to the 3 clock genes Per1, Cry2, and Clock.

```{r}
rhyStats = getRhythmStats(fit, features = clockGeneDt$id, fitType = 'posterior_samples')
rhyStats
```

One use of posterior sampling is to visualize the distribution of each feature's rhythmic statistics post-moderation.

```{r, echo=FALSE, message=FALSE, fig.align='center', fig.width=8, fig.asp=0.5625}
rhyStatsMelt = melt(rhyStats, id.vars = c('feature', 'cond', 'posterior_sample'),
                    variable.name = 'statistic')
rhyStatsMelt = merge(rhyStatsMelt, clockGeneDt, by.x = 'feature', by.y = 'id')

ggplot(data = rhyStatsMelt[!(statistic %in% c('peak_phase', 'trough_phase'))]) +
  geom_density(aes(x = value, fill = cond), alpha = 0.3) +
  facet_grid(rows = vars(sym), cols = vars(statistic), scales = 'free') +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  scale_fill_brewer(palette = 'Dark2')
```

## Get credible intervals for rhythmic statistics

Another reason to use the posterior sampling capability of LimoRhyde2 is to get uncertainty estimates of various quantities of interest. Here, we use `getStatsIntervals` to get a table of upper and lower bounds for each feature-rhythmic-statistic pair.

```{r}
rhyStatsInterval = getStatsIntervals(rhyStats, method = 'eti')
rhyStatsInterval
```

## Get differential rhythmic statistics

We can use posterior sampling to examine differential rhythmicity as well. Here, we produce a table of the upper and lower bounds of the posterior samples of the clock genes.

```{r}
diffRhyStats = getDiffRhythmStats(fit, rhyStats, levels(metadata$cond))
diffRhyStats[clockGeneDt, sym := i.sym, on = .(feature = id)]

diffRhyStatsInterval = getStatsIntervals(diffRhyStats, method = 'eti')
```

We can also see how differentially rhythmic statistics vary within a particular feature by plotting the posterior samples.

```{r, fig.align='center', fig.width=8, fig.asp=0.5625}
ggplot(data = diffRhyStats) +
  geom_point(aes(x = diff_peak_phase, y = diff_peak_trough_amp, color = sym),
             alpha = 0.5) +
  labs(color = 'Feature',
       x = 'Difference in phase (h)',
       y = 'Difference in peak-to-trough amplitude (norm.)') +
  scale_x_continuous(breaks = seq(-12, 12, by = 4)) +
  scale_color_brewer(type = 'qual')

ggplot(data = diffRhyStats) +
  geom_point(aes(x = diff_mean_value, y = diff_peak_trough_amp, color = sym),
             alpha = 0.5) +
  labs(color = 'Feature',
       x = 'Difference in mean expression (norm.)',
       y = 'Difference in peak-to-trough amplitude (norm.)') +
  scale_color_brewer(type = 'qual')


```