---
title: "Quantifying rhythmicity with LimoRhyde2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quantifying-rhythmicity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

We begin by preparing the R environment.

```{r setup}
library(limorhyde2)
library(data.table)
library(ggplot2)

dataPath = '../data/'
```
Next, we load the sample data.

```{r}
load(file.path(dataPath, 'mouse_liver_d.rda'))
load(file.path(dataPath, 'mouse_liver_md.rda'))
```

The first step of the LimoRhyde2 workflow is to obtain a linear model fit using the
`getModelFit` function. For the example data, the name of the column containing 
the time at which a sample was taken is 'time', and in our sample analysis,
we want to fit the data using three offsets.

```{r, results='hide'}
f = getModelFit(y = mouse_liver_d, metadata = mouse_liver_md, nShifts = 3, 
                timeColname = 'time')
```

Next, we obtain a model with moderated estimates using `getPosteriorFit`. We use
the argument `covMethod = 'data-driven'` to fit a model based on covariance matrices
generated from the data.

```{r, results='hide'}
mf = getPosteriorFit(f, covMethod = 'data-driven')
```

From the moderated model, we can get obtain estimates of the rhythmic statistics
for a select set of genes using `getRhythmStats`. Note that `getRhythmStats` can
be used with any `limoRhyde2` fit object, moderated or raw.

```{r}
set.seed(21056)
sampleGenes = sample(rownames(mouse_liver_d), 5)

rs = getRhythmStats(mf, fitType = 'posterior_mean', features = sampleGenes)
```

We can also fit time courses based on the shrunken estimates using `getFittedValues`.

```{r, results='hide'}
fv = getFittedValues(mf, times = seq(0, 24, by = 2), 
                     fitType = 'posterior_mean', features = sampleGenes)
```

Below, we plot the fitted values from the moderated model.

```{r}

ggplot(data = fv, aes(x = time, y = value)) +
  geom_point() +
  facet_wrap(vars(feature), scales = 'free_y')

```