---
title: "Analyzing RNA-seq data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing RNA-seq data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.align = 'center',
  fig.retina = 2,
  eval = FALSE)

foreach::registerDoSEQ()
```

## Introduction

Here we show two options for using `limorhyde2` to analyze RNA-seq data: [limma-voom](https://bioconductor.org/packages/release/bioc/html/limma.html) and [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html). The two approaches give very similar results.

This vignette assumes you are starting with the output of [tximport](https://bioconductor.org/packages/release/bioc/html/tximport.html).

## Load the data

You will need two objects:

1. `txi`, a list from `tximport`
2. `metadata`, a `data.frame` having one row per sample

The rows in `metadata` must correspond to the columns of the elements of `txi`.

```{r}
library('limorhyde2')
# txi = ?
# metadata = ?
```

## Filter out lowly expressed genes

There are many reasonable strategies to do this, here is one.

```{r}
keep = rowSums(edgeR::cpm(txi$counts) >= 0.5) / ncol(txi$counts) >= 0.75

txiKeep = txi
for (name in c('counts', 'length')) {
  txiKeep[[name]] = txi[[name]][keep, ]}
```

## Fill in counts for sample-gene pairs having zero counts

This avoids unrealistically low log2 CPM values and thus artificially inflated effect size estimates.

```{r}
for (i in 1:nrow(txiKeep$counts)) {
  idx = txiKeep$counts[i, ] > 0
  txiKeep$counts[i, !idx] = min(txiKeep$counts[i, idx])}
```

## Option 1: limma-voom

```{r}
y = edgeR::DGEList(txiKeep$counts)
y = edgeR::calcNormFactors(y)

fit = getModelFit(y, metadata, ..., method = 'voom') # replace '...' as appropriate for your data
```

## Option 2: DESeq2

The second and third arguments to `DESeqDataSetFromTxImport()` are required, but will not be used by `limorhyde2`.

```{r}
y = DESeq2::DESeqDataSetFromTximport(txiKeep, metadata, ~1)

fit = getModelFit(y, metadata, ..., method = 'deseq2') # replace '...' as appropriate for your data
```

## Continue using `limorhyde2`

Regardless of which option you choose, the next steps are the same: `getPosteriorFit()`, `getRhythmStats()`, etc.
